Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-06-19 manuscript.Rmd
garment details

abstract: |
  Long form audio recordings have had a transformational effect on the study of infant language acquisition by using mobile, unobtrusive devices to gather full-day, real-time data that can be automatically scored. How can we produce similar data in service of measuring infants' everyday motor behaviors, such as body position? The aim of the current study was to validate long form recordings of infant position (supine, prone, sitting, upright, held by caregiver) based on machine learning classification of data from inertial sensors worn on infants' ankles and hips. Using over 100 hours of video recordings synchronized with inertial sensor data from infants in their homes, we demonstrate that body position classifications are sufficiently accurate to measure infant behavior. Moreover, classification remained accurate when predicting behavior later in the session when infants and caregivers were unsupervised and went about their normal activities, showing that the method can handle the challenge of measuring unconstrained, natural activity. Next, we show that the inertial sensing method has convergent validity by replicating age differences in body position found using other methods with full-day data captured from inertial sensors. We end the paper with a discussion of the novel opportunities that long form motor recordings afford for understanding infant learning and development.

@@

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A *proximal comparison* period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. This "semi-supervised" period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The **first goal** of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using "individual models"---models that were trained on one participant's data to predict their later behavior---compared with a "group model" that aggregated data from all infants to create a single body position classifier [@FranchakScott2021], so we compared both modeling approaches in the current investigation.

@@

Four inertial movement units (IMUs) were used to record infant movement across the day (MC10 Biostamp). A custom garment was designed to hold the IMUs: Internal pockets were sewn into a snug-fitting pair of infant leggings so that IMUs would stay close to the body (reducing vibration), which also prevented infants from pulling out the sensors. A pocket over the hip and a pocket just above the ankle were sewn on the lateral surface of the right and left legs of the garment. Each sensor recorded accelerometer and gyroscope data throughout the day, with sufficient battery and onboard storage to record for approximately 12 hours. We chose a sampling rate of 62.5 Hz (one of the available preset rates) based on prior work that used rates of 50-64 Hz [@FranchakScott2021; @AiraksinenRasanen2020; @YaoPlotz2019]. Infants also wore a LENA&reg; recorder throughout the day in the front pocket of a LENA&reg; shirt, located near the infant’s chest, to determine whether data could be simultaneously recorded from the LENA&reg; and IMU sensors. However, LENA&reg; data were not analyzed as part of the current study.

@@

The same machine learning classification process was used as in prior work [@FranchakScott2021]. Using the synchronization point, human-coded body annotations from video were linked to the corresponding times in the IMU time series data. A single, merged dataset was created with synchronized accelerometer signals (in three orientations: X, Y, and Z) and gyroscope signals (in three orientations: roll, pitch, and yaw) for each of the four sensors (left hip, right hip, left ankle, right ankle) with the corresponding timestamp and body position code using the *timetk* package [@DanchoVaughan2023] and the *lubridate* package [@GrolemundWickham2011] in R version 4.1.2 [@R42].

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification [@FranchakScott2021; @NamPark2013; @AiraksinenRasanen2020]. The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right hip X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for > 75% (3 s) of time within the window to ensure that motion signals could be linked to an unambiguous example of each behavior.

@@

```{r metrics-reduce, cache=TRUE, cache.extra=tools::md5sum("data/reduce/reduce_metrics_right_side.csv")}
all_files = list.files("data/reduce", full.names = T, pattern = ".csv")
overall_files = all_files %>% discard(str_detect(., pattern = "class"))
sensor_comp <- read_csv(overall_files, id = "type")
sensor_comp <- sensor_comp %>% mutate(type = str_remove(type, "data/reduce/reduce_metrics_"),
                              type = str_remove(type, ".csv"),
                              type = factor(type,
                              levels = c("all", "left_side", "right_side", "hips", "ankles",
                                         "left_hip","left_ankle", "right_hip", "right_ankle"),
                              labels = c("All", "Left Hip/Ankle", "Right Hip/Ankle", "Both Hips", "Both Ankles", 
                                         "Left Hip", "Left Ankle", "Right Hip", "Right Ankle")),
                              Model = factor(model, levels = c("group", "split", labels = c("Group", "Individual"))))

@@

apa_table(metrics_sensor, col.names = c("Sensors", "Median", "Mean", "SD", "Median", "Mean", "SD"),    col_spanners = list("Group" = c(2, 4), "Individual" = c(5, 7)), midrules = c(1,5),
          caption = "Median, mean, and SD of overall accuracy calculated with different sets of sensor features. The top row shows performance using all features calculated from the four sensors. Rows 2-5 show accuracy using pairs of sensors (left hip and ankle, right hip and ankle, left and right hip, left and right ankle), and rows 6-9 show accuracy using each individual sensor.")
metrics_sensor <- metrics_sensor %>% column_to_rownames("type")
```

@@

Similar to past work [@AiraksinenRasanen2020], overall accuracy decreased when fewer sensors were used. Table \@ref(tab:metrics-reduce) compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model *M* = `r metrics_sensor["All", "mean_group"]`; individual model *M* = `r metrics_sensor["All", "mean_split"]`), and the lowest when using only a single sensor (left ankle group model *M* = `r metrics_sensor["Left Ankle", "mean_group"]`; right ankle individual model *M* = `r metrics_sensor["Right Ankle", "mean_split"]`). Some sensor pairs showed comparable accuracy to using all four sensors: Left ankle and hip has group model accuracy of *M* = `r metrics_sensor["Left Hip/Ankle", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Left Hip/Ankle", "mean_split"]`, only 2.2% and 1.5% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_split"]`, 12.7% and 7.2% worse than using all four sensors.

@@

As in past work [@AiraksinenRasanen2020], performance was best when using all four sensors and degraded when using any subset of two (i.e., hips only, left leg only). Small decrements and accuracy were seen for some pairs (particularly when using the left ankle and hip), but decrements for other pairs (both ankles) or individual sensors were substantial. Using a hip and ankle pair instead of four sensors would be more cost effective without a large decrease in accuracy, but in some applications the increased accuracy might be worth the cost. Beyond group and individual models and sensor pairings, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we chose to report the best approach, not every possible variation in modeling. However, our openly shared data and code allow researcher to experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different hyperparameter values. We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance was degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no benefit to performance by tailoring to the particular age group.

@@

Second, we relied on caregiver logs of infants' naps and times when sensors were removed to separate usable data (times when infants wore the sensors while awake) from unusable data. Future work should aim to further assess caregivers' perceptions about the usability and comfort of the garment. Although infants wore the sensor garment throughout all of the desired times, we did not collect independent data (aside from caregivers' logs) to verify wear time.  Most caregivers were diligent about completing logs, however, there were a few cases that caregivers may have failed to report naps (younger infants \#1 and \#4). In the future, algorithms can be developed to automatically identify periods of sleep and times when sensors are off the body. Such algorithms exist for adults using wrist-worn sensors, however, we did not use them because they have not been validated for infant participants wearing sensors on hips/ankles. Differences in infants' positioning while asleep (held in caregivers arms, laying in cribs, reclined in strollers and carseats) might make classifying sleep more difficult compared with adults based on movement.

>>>>

abstract: |
  Long form audio recordings have had a transformational effect on the study of infant language acquisition by using mobile, unobtrusive devices to gather full-day, real-time data that can be automatically scored. How can we produce similar data in service of measuring infants' everyday motor behaviors, such as body position? The aim of the current study was to validate long form recordings of infant position (supine, prone, sitting, upright, held by caregiver) based on machine learning classification of data from inertial sensors worn on infants' ankles and thighs. Using over 100 hours of video recordings synchronized with inertial sensor data from infants in their homes, we demonstrate that body position classifications are sufficiently accurate to measure infant behavior. Moreover, classification remained accurate when predicting behavior later in the session when infants and caregivers were unsupervised and went about their normal activities, showing that the method can handle the challenge of measuring unconstrained, natural activity. Next, we show that the inertial sensing method has convergent validity by replicating age differences in body position found using other methods with full-day data captured from inertial sensors. We end the paper with a discussion of the novel opportunities that long form motor recordings afford for understanding infant learning and development.

@@

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each thigh) and a video camera to collect ground truth data about infant body position. A *proximal comparison* period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. This "semi-supervised" period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The **first goal** of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using "individual models"---models that were trained on one participant's data to predict their later behavior---compared with a "group model" that aggregated data from all infants to create a single body position classifier [@FranchakScott2021], so we compared both modeling approaches in the current investigation.

@@

Four inertial movement units (IMUs) were used to record infant movement across the day (MC10 Biostamp). A custom garment was designed to hold the IMUs: Internal pockets were sewn into a snug-fitting pair of infant leggings so that IMUs would stay close to the body (reducing vibration), which also prevented infants from pulling out the sensors. A pocket over the thigh and a pocket just above the ankle were sewn on the lateral surface of the right and left legs of the garment. Multiple sizes of the garment were created (modeled on US 3-6 mo, 6-9 mo, 9-12 mo, and 12-18 mo sizing). Caregivers indicated in advance which size would fit their infant, and to "size down" if on the cusp to ensure a snug fit and reduce sensor movement. Each garment had a distinct pattern on the seat of the pants so that caregivers could identify front versus back and place the garment in the correct orientation.

Each sensor recorded accelerometer and gyroscope data throughout the day, with sufficient battery and onboard storage to record for approximately 12 hours. We chose a sampling rate of 62.5 Hz (one of the available preset rates) based on prior work that used rates of 50-64 Hz [@FranchakScott2021; @AiraksinenRasanen2020; @YaoPlotz2019]. Infants also wore a LENA&reg; recorder throughout the day in the front pocket of a LENA&reg; shirt, located near the infant’s chest, to determine whether data could be simultaneously recorded from the LENA&reg; and IMU sensors. However, LENA&reg; data were not analyzed as part of the current study.

```{r garment, out.width="0.75\\linewidth", include=TRUE, fig.align="center", fig.cap="Sensor garment worn by infant participant. Four IMUs were placed in interior pockets sewn into a tightly-fitting pair of infant leggings. White dashed rectangles mark the approximate locations of each sensor pocket (above the left and right ankles and on the left and right thighs, just below the thighs). A white dashed rectangle also marks the LENA audio recorder worn in a pocket on the infant's shirt.", echo=FALSE}
knitr::include_graphics("figures/garment.jpg")
```

@@

The same machine learning classification process was used as in prior work [@FranchakScott2021]. Using the synchronization point, human-coded body annotations from video were linked to the corresponding times in the IMU time series data. A single, merged dataset was created with synchronized accelerometer signals (in three orientations: X, Y, and Z) and gyroscope signals (in three orientations: roll, pitch, and yaw) for each of the four sensors (left thigh, right thigh, left ankle, right ankle) with the corresponding timestamp and body position code using the *timetk* package [@DanchoVaughan2023] and the *lubridate* package [@GrolemundWickham2011] in R version 4.1.2 [@R42].

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification [@FranchakScott2021; @NamPark2013; @AiraksinenRasanen2020]. The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right thigh X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for > 75% (3 s) of time within the window to ensure that motion signals could be linked to an unambiguous example of each behavior.

@@

```{r metrics-reduce, cache=TRUE, cache.extra=tools::md5sum("data/reduce/reduce_metrics_right_side.csv")}
all_files = list.files("data/reduce", full.names = T, pattern = ".csv")
overall_files = all_files %>% discard(str_detect(., pattern = "class"))
sensor_comp <- read_csv(overall_files, id = "type")
sensor_comp <- sensor_comp %>% mutate(type = str_remove(type, "data/reduce/reduce_metrics_"),
                              type = str_remove(type, ".csv"),
                              type = factor(type,
                              levels = c("all", "left_side", "right_side", "hips", "ankles",
                                         "left_hip","left_ankle", "right_hip", "right_ankle"),
                              labels = c("All", "Left Hip/Ankle", "Right Thigh/Ankle", "Both Thighs", "Both Ankles", 
                                         "Left Thigh", "Left Ankle", "Right Thigh", "Right Ankle")),
                              Model = factor(model, levels = c("group", "split", labels = c("Group", "Individual"))))

@@

apa_table(metrics_sensor, col.names = c("Sensors", "Median", "Mean", "SD", "Median", "Mean", "SD"),    col_spanners = list("Group" = c(2, 4), "Individual" = c(5, 7)), midrules = c(1,5),
          caption = "Median, mean, and SD of overall accuracy calculated with different sets of sensor features. The top row shows performance using all features calculated from the four sensors. Rows 2-5 show accuracy using pairs of sensors (left thigh and ankle, right thigh and ankle, left and right thigh, left and right ankle), and rows 6-9 show accuracy using each individual sensor.")
metrics_sensor <- metrics_sensor %>% column_to_rownames("type")
```

@@

Similar to past work [@AiraksinenRasanen2020], overall accuracy decreased when fewer sensors were used. Table \@ref(tab:metrics-reduce) compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model *M* = `r metrics_sensor["All", "mean_group"]`; individual model *M* = `r metrics_sensor["All", "mean_split"]`), and the lowest when using only a single sensor (left ankle group model *M* = `r metrics_sensor["Left Ankle", "mean_group"]`; right ankle individual model *M* = `r metrics_sensor["Right Ankle", "mean_split"]`). Some sensor pairs showed comparable accuracy to using all four sensors: Left ankle and thigh has group model accuracy of *M* = `r metrics_sensor["Left Thigh/Ankle", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Left Thigh/Ankle", "mean_split"]`, only 2.2% and 1.5% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_split"]`, 12.7% and 7.2% worse than using all four sensors.

@@

As in past work [@AiraksinenRasanen2020], performance was best when using all four sensors and degraded when using any subset of two (i.e., thighs only, left leg only). Small decrements and accuracy were seen for some pairs (particularly when using the left ankle and thigh), but decrements for other pairs (both ankles) or individual sensors were substantial. Using a thigh and ankle pair instead of four sensors would be more cost effective without a large decrease in accuracy, but in some applications the increased accuracy might be worth the cost. Beyond group and individual models and sensor pairings, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we chose to report the best approach, not every possible variation in modeling. However, our openly shared data and code allow researcher to experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different hyperparameter values. We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance was degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no benefit to performance by tailoring to the particular age group.

@@

Second, we relied on caregiver logs of infants' naps and times when sensors were removed to separate usable data (times when infants wore the sensors while awake) from unusable data. Future work should aim to further assess caregivers' perceptions about the usability and comfort of the garment. Although infants wore the sensor garment throughout all of the desired times, we did not collect independent data (aside from caregivers' logs) to verify wear time.  Most caregivers were diligent about completing logs, however, there were a few cases that caregivers may have failed to report naps (younger infants \#1 and \#4). In the future, algorithms can be developed to automatically identify periods of sleep and times when sensors are off the body. Such algorithms exist for adults using wrist-worn sensors, however, we did not use them because they have not been validated for infant participants wearing sensors on thighs/ankles. Differences in infants' positioning while asleep (held in caregivers arms, laying in cribs, reclined in strollers and carseats) might make classifying sleep more difficult compared with adults based on movement.

### 2023-06-19 manuscript.tex
garment details

\abstract{%
Long form audio recordings have had a transformational effect on the study of infant language acquisition by using mobile, unobtrusive devices to gather full-day, real-time data that can be automatically scored. How can we produce similar data in service of measuring infants' everyday motor behaviors, such as body position? The aim of the current study was to validate long form recordings of infant position (supine, prone, sitting, upright, held by caregiver) based on machine learning classification of data from inertial sensors worn on infants' ankles and hips. Using over 100 hours of video recordings synchronized with inertial sensor data from infants in their homes, we demonstrate that body position classifications are sufficiently accurate to measure infant behavior. Moreover, classification remained accurate when predicting behavior later in the session when infants and caregivers were unsupervised and went about their normal activities, showing that the method can handle the challenge of measuring unconstrained, natural activity. Next, we show that the inertial sensing method has convergent validity by replicating age differences in body position found using other methods with full-day data captured from inertial sensors. We end the paper with a discussion of the novel opportunities that long form motor recordings afford for understanding infant learning and development.
}

@@

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A \emph{proximal comparison} period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. This ``semi-supervised'' period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The \textbf{first goal} of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using ``individual models''---models that were trained on one participant's data to predict their later behavior---compared with a ``group model'' that aggregated data from all infants to create a single body position classifier \autocite{FranchakScott2021}, so we compared both modeling approaches in the current investigation.

@@

Four inertial movement units (IMUs) were used to record infant movement across the day (MC10 Biostamp). A custom garment was designed to hold the IMUs: Internal pockets were sewn into a snug-fitting pair of infant leggings so that IMUs would stay close to the body (reducing vibration), which also prevented infants from pulling out the sensors. A pocket over the hip and a pocket just above the ankle were sewn on the lateral surface of the right and left legs of the garment. Each sensor recorded accelerometer and gyroscope data throughout the day, with sufficient battery and onboard storage to record for approximately 12 hours. We chose a sampling rate of 62.5 Hz (one of the available preset rates) based on prior work that used rates of 50-64 Hz \autocite{FranchakScott2021,AiraksinenRasanen2020,YaoPlotz2019}. Infants also wore a LENA® recorder throughout the day in the front pocket of a LENA® shirt, located near the infant's chest, to determine whether data could be simultaneously recorded from the LENA® and IMU sensors. However, LENA® data were not analyzed as part of the current study.

@@

The same machine learning classification process was used as in prior work \autocite{FranchakScott2021}. Using the synchronization point, human-coded body annotations from video were linked to the corresponding times in the IMU time series data. A single, merged dataset was created with synchronized accelerometer signals (in three orientations: X, Y, and Z) and gyroscope signals (in three orientations: roll, pitch, and yaw) for each of the four sensors (left hip, right hip, left ankle, right ankle) with the corresponding timestamp and body position code using the \emph{timetk} package \autocite{DanchoVaughan2023} and the \emph{lubridate} package \autocite{GrolemundWickham2011} in R version 4.1.2 \autocite{R42}.

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification \autocite{FranchakScott2021,NamPark2013,AiraksinenRasanen2020}. The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right hip X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for \textgreater{} 75\% (3 s) of time within the window to ensure that motion signals could be linked to an unambiguous example of each behavior.

@@

@@

@@

Beyond group and individual models, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we chose to report the best approach, not every possible variation in modeling. However, our openly shared data and code allow researcher to experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different hyperparameter values. As in past work \autocite{AiraksinenRasanen2020}, performance was best when using all four sensors and degraded when using any subset of two (i.e., hips only, left leg only). We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance was degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no benefit to performance by tailoring to the particular age group.

>>>>

\abstract{%
Long form audio recordings have had a transformational effect on the study of infant language acquisition by using mobile, unobtrusive devices to gather full-day, real-time data that can be automatically scored. How can we produce similar data in service of measuring infants' everyday motor behaviors, such as body position? The aim of the current study was to validate long form recordings of infant position (supine, prone, sitting, upright, held by caregiver) based on machine learning classification of data from inertial sensors worn on infants' ankles and thighs. Using over 100 hours of video recordings synchronized with inertial sensor data from infants in their homes, we demonstrate that body position classifications are sufficiently accurate to measure infant behavior. Moreover, classification remained accurate when predicting behavior later in the session when infants and caregivers were unsupervised and went about their normal activities, showing that the method can handle the challenge of measuring unconstrained, natural activity. Next, we show that the inertial sensing method has convergent validity by replicating age differences in body position found using other methods with full-day data captured from inertial sensors. We end the paper with a discussion of the novel opportunities that long form motor recordings afford for understanding infant learning and development.
}

@@

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on 34 testing sessions from 22 infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each thigh) and a video camera to collect ground truth data about infant body position. A \emph{proximal comparison} period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. This ``semi-supervised'' period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play while they received instructions from the experimenter. The \textbf{first goal} of the current study was to determine the accuracy of body position classification during the proximal comparison period using this novel, semi-supervised procedure in participants' homes. Past work found better performance using ``individual models''---models that were trained on one participant's data to predict their later behavior---compared with a ``group model'' that aggregated data from all infants to create a single body position classifier \autocite{FranchakScott2021}, so we compared both modeling approaches in the current investigation.

@@

Four inertial movement units (IMUs) were used to record infant movement across the day (MC10 Biostamp). A custom garment was designed to hold the IMUs: Internal pockets were sewn into a snug-fitting pair of infant leggings so that IMUs would stay close to the body (reducing vibration), which also prevented infants from pulling out the sensors. A pocket over the thigh and a pocket just above the ankle were sewn on the lateral surface of the right and left legs of the garment. Multiple sizes of the garment were created (modeled on US 3-6 mo, 6-9 mo, 9-12 mo, and 12-18 mo sizing). Caregivers indicated in advance which size would fit their infant, and to ``size down'' if on the cusp to ensure a snug fit and reduce sensor movement. Each garment had a distinct pattern on the seat of the pants so that caregivers could identify front versus back and place the garment in the correct orientation.

Each sensor recorded accelerometer and gyroscope data throughout the day, with sufficient battery and onboard storage to record for approximately 12 hours. We chose a sampling rate of 62.5 Hz (one of the available preset rates) based on prior work that used rates of 50-64 Hz \autocite{FranchakScott2021,AiraksinenRasanen2020,YaoPlotz2019}. Infants also wore a LENA® recorder throughout the day in the front pocket of a LENA® shirt, located near the infant's chest, to determine whether data could be simultaneously recorded from the LENA® and IMU sensors. However, LENA® data were not analyzed as part of the current study.

{\centering \includegraphics[width=0.75\linewidth]{figures/garment}

\caption{Sensor garment worn by infant participant. Four IMUs were placed in interior pockets sewn into a tightly-fitting pair of infant leggings. White dashed rectangles mark the approximate locations of each sensor pocket (above the left and right ankles and on the left and right thighs, just below the thighs). A white dashed rectangle also marks the LENA audio recorder worn in a pocket on the infant's shirt.}\label{fig:garment}
\end{figure}

@@

The same machine learning classification process was used as in prior work \autocite{FranchakScott2021}. Using the synchronization point, human-coded body annotations from video were linked to the corresponding times in the IMU time series data. A single, merged dataset was created with synchronized accelerometer signals (in three orientations: X, Y, and Z) and gyroscope signals (in three orientations: roll, pitch, and yaw) for each of the four sensors (left thigh, right thigh, left ankle, right ankle) with the corresponding timestamp and body position code using the \emph{timetk} package \autocite{DanchoVaughan2023} and the \emph{lubridate} package \autocite{GrolemundWickham2011} in R version 4.1.2 \autocite{R42}.

Classification training and prediction was conducted on a windowed dataset that summarized the raw, 62.5 Hz motion signals within 4-s windows. Overlapping moving windows captured 4-s (250 samples) of data starting each second, which is a common unit of analysis in prior studies of human activity classification \autocite{FranchakScott2021,NamPark2013,AiraksinenRasanen2020}. The 250 samples of data in each window were aggregated into a variety of motion features---summary statistics that could be fed into the machine learning model. The minimum, maximum, 25th percentile, 75th percentile, mean, median, skew, kurtosis, standard deviation, and sum were computed for each signal (e.g., right thigh X linear acceleration, left ankle pitch angular acceleration). The 10 summary statistics and 24 sensor signals generated 240 columns of motion features that described movement within each window. Furthermore, a series of cross-sensor and cross-orientation summaries (such as the correlation, magnitude, and difference between pairs of sensors) added an additional 196 columns of motion features. The 436 total motion features corresponded to a single body annotation code for each 4-s window. Windows were only used for training/testing if they contained a single body position for \textgreater{} 75\% (3 s) of time within the window to ensure that motion signals could be linked to an unambiguous example of each behavior.

@@

\caption{\label{tab:metrics-reduce}Median, mean, and SD of overall accuracy calculated with different sets of sensor features. The top row shows performance using all features calculated from the four sensors. Rows 2-5 show accuracy using pairs of sensors (left thigh and ankle, right thigh and ankle, left and right thigh, left and right ankle), and rows 6-9 show accuracy using each individual sensor.}

\begin{tabular}{lllllll}
\toprule
 & \multicolumn{3}{c}{Group} & \multicolumn{3}{c}{Individual} \\
\cmidrule(r){2-4} \cmidrule(r){5-7}
Sensors & Median & Mean & SD & Median & Mean & SD\\
\midrule
All & 0.894 & 0.846 & 0.131 & 0.933 & 0.916 & 0.072\\ \midrule
Left Hip/Ankle & 0.891 & 0.824 & 0.177 & 0.915 & 0.901 & 0.071\\
Right Thigh/Ankle & 0.842 & 0.800 & 0.167 & 0.897 & 0.898 & 0.086\\
Both Thighs & 0.859 & 0.791 & 0.195 & 0.907 & 0.898 & 0.072\\
Both Ankles & 0.797 & 0.719 & 0.196 & 0.864 & 0.844 & 0.108\\ \midrule
Left Thigh & 0.842 & 0.762 & 0.208 & 0.889 & 0.872 & 0.077\\
Left Ankle & 0.717 & 0.667 & 0.208 & 0.851 & 0.825 & 0.117\\
Right Thigh & 0.846 & 0.772 & 0.178 & 0.896 & 0.876 & 0.092\\
Right Ankle & 0.704 & 0.683 & 0.160 & 0.830 & 0.819 & 0.114\\
\bottomrule
\end{tabular}

@@

Similar to past work \autocite{AiraksinenRasanen2020}, overall accuracy decreased when fewer sensors were used. Table \ref{tab:metrics-reduce} compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model \emph{M} = 0.846; individual model \emph{M} = 0.916), and the lowest when using only a single sensor (left ankle group model \emph{M} = 0.667; right ankle individual model \emph{M} = 0.819). Some sensor pairs showed comparable accuracy to using all four sensors: Left ankle and thigh has group model accuracy of \emph{M} = NA and individual model accuracy of \emph{M} = NA, only 2.2\% and 1.5\% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of \emph{M} = 0.719 and individual model accuracy of \emph{M} = 0.844, 12.7\% and 7.2\% worse than using all four sensors.

@@

As in past work \autocite{AiraksinenRasanen2020}, performance was best when using all four sensors and degraded when using any subset of two (i.e., thighs only, left leg only). Small decrements and accuracy were seen for some pairs (particularly when using the left ankle and thigh), but decrements for other pairs (both ankles) or individual sensors were substantial. Using a thigh and ankle pair instead of four sensors would be more cost effective without a large decrease in accuracy, but in some applications the increased accuracy might be worth the cost. Beyond group and individual models and sensor pairings, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we chose to report the best approach, not every possible variation in modeling. However, our openly shared data and code allow researcher to experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different hyperparameter values. We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance was degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no benefit to performance by tailoring to the particular age group.
